Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-21 ms.qmd
Fixed up some text

With acclimation time and maximum temperature as moderators in all our models we proceeded to build separate models that tested our core questions. All estimates from our models are therefore conditioned on an average acclimation time (i.e.,
`r mean(data_wide$acclim_period, na.rm = TRUE)` days) and maximum temperature (i.e., 23°C) across the dataset.

@@

Finally, we tested whether the change in $lnRR_{Q_{10_{acclimation}}}$ and $lnVR_{Q_{10}}$ were predicted by climate variability (CV). We only used $lnRR_{Q_{10_{acclimation}}}$ for these models because our predictions were specifically focused on acclimation responses. We used all $lnVR_{Q_{10}}$ data because there were no differences between $lnVR_{Q_{10_{acclimation}}}$ and $lnVR_{Q_{10_{acute}}}$ and we had no *a priori* reason to believe climate would impact acclimation and acute $lnVR_{Q_{10}}$ differently. We fit models that included an interaction between habitat type and thermal coefficient of variability (CV) as moderators (referred to as "Model 5"). We also explored whether environmental predictability explained capacity for acclimation; we estimated predictability as the correlation of temperatures across months at a given location. However, such analyses are challenging to interpret because the temporal scale that is biologically relevant to different organisms will be different making the choice of lag to estimate the correlation difficult to apply across taxa. As such, we report a simple analysis in the *Supplementary Materials* but note they do not differ from our CV analysis.

@@

We conducted a simple simulation as a sensitivity analysis to better understand the characteristics of performance curves that could lead to our observed changes in variance across temperatures. The simulation varied among-individual variation in performance curves to identify the parameters that could produce the results we observed. To simulate performance curves, we used an asymmetrical Gaussian function [@angilletta2009thermal]:

@@

Physiological rates increased more in terrestrial ectotherms ($\mu$ = `r mean(t_RR_mean)` 95%CI: `r quantile(t_RR_mean, 0.025)` to `r quantile(t_RR_mean, 0.975)`) compared to marine ($\mu$ = `r mean(m_RR_mean)` 95%CI: `r quantile(m_RR_mean, 0.025)` to `r quantile(m_RR_mean, 0.975)`) and freshwater ectotherms ($\mu$ = `r mean(f_RR_mean)` 95%CI: `r quantile(f_RR_mean, 0.025)` to `r quantile(f_RR_mean, 0.975)`), but did not differ significantly betweem aquatic and terrestrial habiatats (Terrestrial - Marine: $\beta$ = `r mean(t_m_rates)` 95%CI: `r quantile(t_m_rates, 0.025)` to `r quantile(t_m_rates, 0.975)`, $p_{MCMC}$ = `p_value(pmcmc(t_m_rates))`; Terrestrial - Freshwater: $\beta$ = `r mean(t_f_rates)` 95%CI: `r quantile(t_f_rates, 0.025)` to `r quantile(t_f_rates, 0.975)`, $p_{MCMC}$ = `p_value(pmcmc(t_f_rates))`) ("Model 2"). However, capacity for acclimation depended on the habitat. Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-2]A) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r abs(mean(change_f))`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r abs(mean(change_m))`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation with a `r abs(mean(change_t))`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-2]A).

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B). Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms,  a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms and a `r mean(f_VR_mean_ef)`% (95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))`) increase in variance for freshwater ectotherms across 10°C (@fig-2; results from "Model 2"). Analysis of $lnCVR_{Q_{10}}$, which accounts for changes in mean physiological rates, also showed that the relative variance for terrestrial ectotherms decreased compared to marine and freshwater ectotherms, suggesting that increases in variance are less than expected for ectotherms occupying terrestrial habitats (See *Supplementary Materials*).

Physiological rate variance increased significantly more in freshwater compared to terrestrial ectotherms for acute responses ($\beta$ = `r mean(f_t_c_a)`, 95% CI: `r quantile(f_t_c_a, 0.025)` to `r quantile(f_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_a))`), but not for acclimation responses because increases in rates were dampened by acclimation resulting in smaller increases in variance ($\beta$ = `r mean(f_t_c_aclim)`, 95% CI: `r quantile(f_t_c_aclim, 0.025)` to `r quantile(f_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_aclim))`). While marine ectotherms had larger increases in variance compared to terrestrial ectotherms these were not significant (Acute: $\beta$ = `r mean(m_t_c_a)`, 95% CI: `r quantile(m_t_c_a, 0.025)` to `r quantile(m_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_a))`; Acclimation: $\beta$ = `r mean(m_t_c_aclim)`, 95% CI: `r quantile(m_t_c_aclim, 0.025)` to `r quantile(m_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_aclim))`)([@fig-2]B). Marine and freshwater habitats did not differ in the extent of variance increases at higher temperatures (Acute: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`; Acclimation: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). There were no differences between $lnVR_{Q_{10_{acute}}}$ and $lnVR_{Q_{10_{acclimation}}}$ within any habitat ([@fig-2]B). Differences in variance between habitats were driven by different scaling relationships between mean and variance with slopes varying across habitat types (see *Supplementary Materials* scaling analyses, @tbl-slopes & @fig-mv, and analyses with $lnCVR_{Q_{10}}$, @fig-s1).

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can; @Sanderson2023; @urban2023and]. Here we show that both acclimation responses ($lnRR_{Q_{10}}$) and increases in physiological rate variance ($lnVR_{Q_{10}}$) of ectotherms varied across habitats. Our results uncover an hitherto unrecognised dynamic where the limited benefits of acclimation are potentially compensated for by increases in trait variance that may have impacts on how ectotherm populations across these different habitats will be able to adapt to increased temperatures.

@@

We show that the capacity for acclimation of physiological rates differs across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. These consistent results are interesting given the different physiological traits measured in these meta-analyses (i.e., thermal limits versus physiological rates).

The change in acclimation $Q_{10}$ we found with our expanded dataset (34% increase in effects) was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms [decrease of 16% versus ~10% in @Seebacher2015-wi], and opposite in terrestrial ectotherms [increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi], possibly due to the use of newly derived $Q_{10}$ effect sizes that allow us to control for sampling variance. Greater capacity for acclimation in aquatic organisms may be the result of fewer opportunities for behavioural thermoregulation in aquatic environments making physiological remodeling important for maintaining homeostasis [@Gunderson2015-hz; @morley2019physiological]. Importantly, even though marine and freshwater ectotherms were capable of partial acclimation, on average, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6), suggesting acclimation will provide limited scope for aquatic ectotherms to adjust their physiology to higher temperatures.

Biochemical traits (e.g., enzyme activities) were more likely to acclimate compared to whole-organisms traits as predicted, but only for marine ectotherms, presumably because of the less complicated physiological networks that underlie specific biochemical processes [@schulte2011thermal]. However, this prediction did not hold up generally with both whole-organism and biochemical traits being equally likely to acclimate in freshwater ectotherms, which may be due to the different types of biochemical processes measured in the two habitats.

### ***Increased variability in physiological rates across habitats: insights into the adaptive potential of physiological processes in the face of climate change?***

Contrary to acclimation capacity variance in physiological rates increased across habitats with effect sizes being 3-5x larger than those observed for acclimation. Mechanistically, it is unclear what exactly is contributing to the increased variation in physiological rates at higher temperatures, but it is likely the result of increased among-individual variability in how biochemical, cellular and physiological processes function at higher temperatures to maintain homeostasis [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins; @fields2001protein; @schulte2011thermal]. Higher temperatures increase membrane fluidity affecting electrochemical gradients and impacting protein structure and function [@somero1995proteins; @fields2001protein; @tattersall2012coping]. Such challenges (among others) may expose among-individual variation within a population. Indeed, even acclimation itself can impact among-individual variation, possibly through changes in individual performance curves [@schulte2011thermal]. Indeed, we find evidence to that effect given that biochemical processes in marine organisms were both more likely to acclimate and have reduced variance increases compared to acute responses. However, this was not a general phenomena given that freshwater ectotherms were able to maintain high variance despite being able to acclimate.

Importantly, increased variance in physiological rates was not equal among terrestrial, marine and freshwater ectotherms, with increases in variance being higher in freshwater ectotherms (~60% increase / 10°C) compared to terrestrial ectotherms (~36% increase / 10°C). One possible hypothesis for the differences in variability we observed across habitats could be that among-individual variation in key parameters affecting the shape of thermal performance curves differ [@angilletta2009thermal; @huey1989evolution; @tattersall2012coping; @rezende2019thermal]. Our simulations suggest that theoretical and observed $lnVR_{Q_{10}}$ match when thermal performance curves have different among-individual variance in thermal optima and breadth across habitats making this hypothesis plausible. Such patterns across habitats are expected given that terrestrial ectotherms should be adapted to warmer and more variable environments. Theoretical models also suggest that populations with greater temporal environmental variability exhibit greater thermal breadth [@lynch1987environmental]. While we did not find support that thermal variation co-varied with $lnVR_{Q_{10}}$ (see below), as would be expected, such analyses will depend on an appropriate measure of temporal variation in temperature that is biologically relevant -- a challenging feat across diverse taxa, but worthy of future investigation.

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change. Assuming changes in variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, smaller increases in physiological variance could limit adaptation in terrestrial habitats more than aquatic habitats in the future [@urban2023and; @hoffmann2011climate]. For example, under climate change we expect an increase in variance in physiological rates of only ~13% in terrestrial habitats whereas for freshwater habitats we expect variation in physiological rates to increase by ~30%. There will obviously be limits to variance increases, and we predict that organisms closer to their upper thermal limits ($CT_{max}$) will have lower $lnVR_{Q_{10}}$ values compared to those farther away from $CT_{max}$. Some evidence points to possible differences across habitats in upper thermal limits already [@Gunderson2015-hz; @Pinsky2019-fn] making this a fruitful future question to explore.

@@

Acclimation capacities are expected to differ between life-stages because of distinct patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022; @sinclair2016can]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], and such costs can be magnified in late life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constrain plasticity.

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may be because both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. A focus on collecting more detailed information on behaviour, dispersal and thermal environments experienced by different life stages is likely to provide a more complete picture on when plasticity differs. To date, however, there are few studies that have focused on how variability in responses change. We would also encourage more empirical focus on this question given its potential ecological and evolutionary implications.

### ***No signature of past climate on capacity for physiological plasticity or changes in variance***

Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when fluctuations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] might suggest that plasticity is more likely to evolve in terrestrial environments. However, if thermal variability is too high and unpredictable, the rates of acclimation slow, and there are costs associated with re-modelling physiological processes [which seem likely, @angilletta2009thermal] then we would instead expect canalised responses [@angilletta2009thermal; @Seebacher2015-wi; @loughland2020differences; @Leung2020; @rescan2022experimental; @leung2023phenotypic]. The lack of acclimation in terrestrial ectotherms we observed is consistent with the latter hypothesis, and is supported by other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz] suggesting that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@hendry2016key; @Leung2020].

Whether population capacity for acclimation is related to the thermal variability (or predictability) it experiences is equivocal. We show no relationship between acclimation capacity and thermal variability in marine, freshwater and terrestrial habitats. Our results are consistent with @Gunderson2015-hz who show no relationship between plasticity in heat tolerance and latitude or thermal seasonality. However, other analyses on heat tolerance limits have found relationships between latitude (a proxy for seasonality) [@morley2019physiological] or even direct measures of thermal variability [@verberk2024heat]. @Seebacher2015-wi also found evidence that acclimation capacity was related to a populations thermal variability, however, relationships depended on the habitat and traits in question. Discrepancies across studies could be related to the taxa included in analyses [e.g., @morley2019physiological], different traits or possibly the fact that different climate projections/models are being used to quantify thermal variability. Modelling realistic microenvironments across such diverse taxa is also challenging because it is unclear what the most appropriate spatial and temporal scale might be that is of evolutionary relevance. Historical temperature time series' may not be representative of the selective environment a population has experienced making relationships between capacity for acclimation and temperature variability (or predictability) difficult to pin down.

@@

It is clear that many fascinating questions remain unanswered that will require greater focus on the consequences of
changes in variance (rather than just the mean). Particularly interesting questions include: How do differences in physiological rate
variance change energy flow across trophic levels within communities? What are the biochemical, cellular, and physiological mechanisms that underlie differences in physiological rate variance across habitats? Are changes in variance in one trait associated with changes in other traits, or do some traits increase while others decrease? Are changes in physiological rate variance correlated with changes in genetic variation? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across major ecosystems globally.

@@

DWAN is supported by an Australian Research Council (ARC) Future Fellowship (FT220100276) and ARC Discovery Project (DP210101152). SN is supported by ARC Discovery Project Grants (DP210100812 and DP230101248).

>>>>

With acclimation time and maximum temperature as moderators in all our models we proceeded to build separate models that tested our core questions. All estimates from our models are therefore conditioned on an average acclimation time (i.e., `r mean(data_wide$acclim_period, na.rm = TRUE)` days) and an average maximum temperature (i.e., 23°C) across the dataset.

@@

Finally, we tested whether the change in $lnRR_{Q_{10_{acclimation}}}$ and $lnVR_{Q_{10}}$ were predicted by climate variability (CV) (details on climate data can be found in the *Supplementary Materials*). We only used $lnRR_{Q_{10_{acclimation}}}$ and $lnVR_{Q_{10_{acclimation}}}$ for these models because our predictions were specifically focused on acclimation responses; though there were no differences between $lnVR_{Q_{10_{acclimation}}}$ and $lnVR_{Q_{10_{acute}}}$. We fit models that included an interaction between habitat type and thermal coefficient of variability (CV) as moderators (referred to as "Model 5"). We also explored whether environmental predictability explained capacity for acclimation; we estimated predictability as the correlation of temperatures across months at a given location. However, such analyses are challenging to interpret because the temporal scale that is biologically relevant to different organisms will be different making the choice of lag to estimate the correlation difficult to apply across taxa. As such, we report a simple analysis in the *Supplementary Materials* but note that it does not differ from our CV analysis.

@@

We conducted a simple simulation as a sensitivity analysis to better understand the characteristics of performance curves that could lead to our observed changes in variance across temperatures and habitats. The simulation varied among-individual variation in performance curves to identify the parameters that could produce the results we observed. To simulate performance curves, we used an asymmetrical Gaussian function [@angilletta2009thermal]:

@@

Physiological rates increased more in terrestrial ectotherms ($\mu$ = `r mean(t_RR_mean)`, 95% CI: `r quantile(t_RR_mean, 0.025)` to `r quantile(t_RR_mean, 0.975)`) compared to marine ($\mu$ = `r mean(m_RR_mean)`, 95% CI: `r quantile(m_RR_mean, 0.025)` to `r quantile(m_RR_mean, 0.975)`) and freshwater ectotherms ($\mu$ = `r mean(f_RR_mean)`, 95% CI: `r quantile(f_RR_mean, 0.025)` to `r quantile(f_RR_mean, 0.975)`), but did not differ significantly betweem aquatic and terrestrial habiatats (Terrestrial - Marine: $\beta$ = `r mean(t_m_rates)`, 95% CI: `r quantile(t_m_rates, 0.025)` to `r quantile(t_m_rates, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(t_m_rates))`; Terrestrial - Freshwater: $\beta$ = `r mean(t_f_rates)`, 95% CI: `r quantile(t_f_rates, 0.025)` to `r quantile(t_f_rates, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(t_f_rates))`) ("Model 2"). However, capacity for acclimation depended on the habitat. Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-2]A) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r abs(mean(change_f))`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r abs(mean(change_m))`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation with a `r abs(mean(change_t))`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-2]A).

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B). Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms,  a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms and a `r mean(f_VR_mean_ef)`% (95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))`) increase in variance for freshwater ectotherms across 10°C (@fig-2; results from "Model 2"). Analysis of $lnCVR_{Q_{10}}$, which accounts for changes in mean physiological rates, also showed that the relative variance for terrestrial ectotherms decreased compared to marine and freshwater ectotherms, suggesting that increases in variance are less than expected for ectotherms occupying terrestrial habitats (See *Supplementary Materials*, @fig-s1).

Physiological rate variance increased significantly more in freshwater compared to terrestrial ectotherms for acute responses ($\beta$ = `r mean(f_t_c_a)`, 95% CI: `r quantile(f_t_c_a, 0.025)` to `r quantile(f_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_a))`), but not for acclimation responses because increases in rates were dampened by acclimation resulting in smaller increases in variance ($\beta$ = `r mean(f_t_c_aclim)`, 95% CI: `r quantile(f_t_c_aclim, 0.025)` to `r quantile(f_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_aclim))`). While marine ectotherms had larger increases in variance compared to terrestrial ectotherms these were not significant (Acute: $\beta$ = `r mean(m_t_c_a)`, 95% CI: `r quantile(m_t_c_a, 0.025)` to `r quantile(m_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_a))`; Acclimation: $\beta$ = `r mean(m_t_c_aclim)`, 95% CI: `r quantile(m_t_c_aclim, 0.025)` to `r quantile(m_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_aclim))`)([@fig-2]B). Marine and freshwater habitats did not differ in the extent of variance increases at higher temperatures (Acute: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`; Acclimation: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). There were no differences between $lnVR_{Q_{10_{acute}}}$ and $lnVR_{Q_{10_{acclimation}}}$ within any habitat ([@fig-2]B). Differences in variance between habitats were driven by different scaling relationships between mean and variance with slopes varying across habitat types (see *Supplementary Materials* scaling analyses, @tbl-slopes & @fig-mv).

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can; @Sanderson2023; @urban2023and]. Here we show that both acclimation responses ($lnRR_{Q_{10}}$) and increases in physiological rate variance ($lnVR_{Q_{10}}$) of ectotherms varied across habitats. Our results uncover an hitherto unrecognised dynamic where the benefits of partial acclimation are paralleled by increases in trait variance that may have impacts on how ectotherm populations across different habitats will be able to adapt to increased temperatures.

@@

We show that the capacity for acclimation of physiological rates differs across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis confirms findings by @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. These consistent results are interesting given the different physiological traits measured in these meta-analyses (i.e., thermal limits versus physiological rates).

The change in acclimation $Q_{10}$ we found in our expanded dataset was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms [decrease of 16% versus ~10% in @Seebacher2015-wi], and opposite in terrestrial ectotherms [increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi]. The difference observed in terrestrial ectotherms between studies may be due to additional data from terrestrial species added in our analysis, and to the use of newly derived $Q_{10}$ effect sizes that allowed us to control for sampling variance. Greater capacity for acclimation in aquatic organisms may be the result of fewer opportunities for behavioural thermoregulation in aquatic environments making physiological remodeling important for maintaining homeostasis [@Gunderson2015-hz; @morley2019physiological]. Importantly, even though marine and freshwater ectotherms were capable of partial acclimation, on average, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6), suggesting that acclimation provides limited scope for aquatic ectotherms to adjust their physiology to higher temperatures.

Biochemical traits (e.g., enzyme activities) were more likely to acclimate compared to whole-organism traits as predicted, but only for marine ectotherms, presumably because of the less complicated physiological networks that underlie specific biochemical processes [@schulte2011thermal]. However, whole-organism and biochemical traits acclimated to the same extent in freshwater ectotherms. Differences between marine and freshwater ectotherms do not appear to be related to the types of traits measured across the habitats but rather that the traits responded differently. However, it is important to note that whole-organism traits in marine ectotherms did respond in the same direction and of a similar magnitude to freshwater ectotherms, so it is possible the difference in marine habitats is simply a sampling artefact.

### ***Increased variability in physiological rates across habitats: adaptive potential of physiological processes in the face of climate change?***

Contrary to acclimation capacity variance in physiological rates increased across habitats with effect sizes being 3-5 times larger than those observed for acclimation of mean trait values. Mechanistically, it is unclear what exactly is contributing to the increased variation in physiological rates at higher temperatures, but it is likely the result of increased among-individual variability in how biochemical, cellular and physiological processes function at higher temperatures to maintain homeostasis [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins; @fields2001protein; @schulte2011thermal]. Higher temperatures increase membrane fluidity affecting electrochemical gradients and impacting protein structure and function [@somero1995proteins; @fields2001protein; @tattersall2012coping]. Such challenges (among others) may expose among-individual variation within a population. Indeed, there is considerable variation in acclimation capacity among individuals which would increase variance in thermal performance curves within populations [@loughland2020differences; @schulte2011thermal].

Importantly, increased variance in physiological rates was not equal among terrestrial, marine and freshwater ectotherms, with increases in variance being higher in freshwater ectotherms (~60% increase / 10°C) compared to terrestrial ectotherms (~36% increase / 10°C). One possible hypothesis for the differences in variability we observed across habitats could be that among-individual variation in key parameters affecting the shape of thermal performance curves differ between habitats [@angilletta2009thermal; @huey1989evolution; @tattersall2012coping; @rezende2019thermal]. Our simulations suggest that theoretical and observed $lnVR_{Q_{10}}$ match when thermal performance curves have different among-individual variance in thermal optima and breadth across habitats making this hypothesis plausible. Such patterns across habitats are expected given that terrestrial ectotherms should be adapted to more extreme and variable thermal environments. Theoretical models also suggest that populations with greater temporal environmental variability exhibit greater thermal breadth [@lynch1987environmental]. However, we did not find support that thermal variation co-varied with $lnVR_{Q_{10}}$ (see below), as would be expected. The relevance of analyses of thermal variability will depend on temporal variation in temperature that is biologically relevant – a challenging feat across diverse taxa, but worthy of future investigation.

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change. Assuming that changes in variation in physiological rates are underpinned by genetic variation, and that there is a genetic correlation with fitness, smaller increases in physiological variance could limit adaptation in terrestrial habitats more than aquatic habitats in the future [@urban2023and; @hoffmann2011climate]. For example, under climate change we expect an increase in variance in physiological rates of only ~13% in terrestrial habitats whereas for freshwater habitats we expect variation in physiological rates to increase by ~30%. There will obviously be limits to variance increases, and we predict that organisms closer to their upper thermal limits ($CT_{max}$) will have lower $lnVR_{Q_{10}}$ values compared to those farther away from $CT_{max}$. Some evidence points to possible differences across habitats in upper thermal limits already [@Gunderson2015-hz; @Pinsky2019-fn], making this a fruitful future question to explore.

@@

Acclimation capacities are expected to differ between life-stages because of distinct patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022; @sinclair2016can]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], such costs can be magnified in later life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constrain plasticity.

Despite these expectations, our analysis does not show any significant differences between early and late life acclimation capacities and little change in the variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait-dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may be because both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar.  A focus on collecting more detailed information on behaviour, dispersal and thermal environments experienced by different life stages is likely to provide a more complete picture on when plasticity differs. We would also encourage more empirical focus on this question and its potential ecological and evolutionary implications.

### ***Past climate does not influence capacity for physiological acclimation or changes in variance***

Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when fluctuations are predictable over time to make environmental cues reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] therefore suggest that plasticity is more likely to evolve in terrestrial environments. However, if thermal variability is too high and unpredictable, the rates of acclimation decrease and there are increased costs associated with re-modelling physiological processes [@angilletta2009thermal] it would instead be expected that phenotypes are canalised during development [@angilletta2009thermal; @Seebacher2015-wi; @loughland2020differences; @Leung2020; @rescan2022experimental; @leung2023phenotypic]. The lack of acclimation in terrestrial ectotherms we observed is consistent with the latter hypothesis, and is supported by other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz] suggesting that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@hendry2016key; @Leung2020].

Whether population capacity for acclimation is related to the thermal variability (or predictability) it experiences is equivocal. We show no relationship between acclimation capacity and thermal variability in marine, freshwater and terrestrial habitats. Our results are consistent with @Gunderson2015-hz who show no relationship between plasticity in heat tolerance and latitude or thermal seasonality. However, other analyses on heat tolerance limits have found relationships between latitude (a proxy for seasonality) [@morley2019physiological] or even direct measures of thermal variability [@verberk2024heat]. @Seebacher2015-wi also found that acclimation capacity was related to a populations thermal variability, however, relationships depended on the habitat and traits in question, and tropical animals showed greater acclimation capacity. Discrepancies across studies could be related to the taxa included in analyses [e.g., @morley2019physiological], different traits or possibly the fact that different climate projections/models are being used to quantify thermal variability. Latitude covaries with a diversity of different ecological attributes aside from temperature [@louthan2021climate], which means it may be capturing other aspects of the environment that affect acclimation capacity. In addition, modelling realistic microenvironments across such diverse taxa is also challenging because it is unclear what the most appropriate spatial and temporal scale might be that is of evolutionary relevance. Historical temperature time series' may not be representative of the selective environment a population has experienced making relationships between capacity for acclimation and temperature variability (or predictability) difficult to pin down.

@@

Many fascinating questions remain unanswered that will require greater focus on the consequences of
changes in variance (rather than just the mean). Particularly interesting questions include: How do differences in physiological rate variance change energy flow across trophic levels within communities? What are the biochemical, cellular, and physiological mechanisms that underlie differences in physiological rate variance across habitats? Are changes in variance in one trait associated with changes in other traits, or do some traits increase while others decrease? Are changes in physiological rate variance correlated with changes in genetic variation? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across major ecosystems globally.

@@

DWAN is supported by an Australian Research Council (ARC) Future Fellowship (FT220100276) and ARC Discovery Project DP210101152. SN is supported by ARC Discovery Project Grants DP210100812 and DP230101248. FS is supported by ARC Discovery Project Grant DP220101342.
